---
title: "transportation model"
author: "sindhuja manda"
date: "2025-10-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA)
```


```{r}
library(knitr)
library(lpSolve)
```

The Transportation model 
problem statement: 

```{r}
# Create the main data frame
table_data <- data.frame(
  Dest1 = c(5, 3, 6, 3, 80),
  Dest2 = c(6, 5, 2, 4, 90),
  Dest3 = c(12, 10, 6, 10, 70),
  ProductionCost = c(300, 200, 250, 150, NA),
  ProductionCapacity = c(100, 75, 85, 40, NA)
)

# Assign row names for sources and demand
rownames(table_data) <- c("Plant1", "Plant2", "Plant3", "Plant4", "MonthlyDemand")

# View the final table
table_data

```

UNBALANCED: 

since the total supply (100 + 75 + 85 + 40 = 300) exceeds total demand (80 + 90 + 70 = 240) by 60 units, we add a dummy destination (a sink) Dest 4 with:

Zero shipping cost from all plants (because it's a dummy)

Demand = 60

```{r}
# Add Dummy Destination (Dest4)
table_data$Dest4 <- c(0, 0, 0, 0, 60)  # zero cost for all plants, demand = 60

# Reorder columns to keep a clean layout
table_data <- table_data[, c("Dest1", "Dest2", "Dest3", "Dest4", "ProductionCost", "ProductionCapacity")]

# View final table
table_data
```

this balances the problem be total demand = 80+90+70+60(dummy) = 300 = total supply.


## Q1: FORMULATION OF TRANSPORTATION MODEL


A) Decision variables

Let
i = 1,2,3,4 index the plants (Plant1..Plant4)

j = 1,2,3,4 index the destinations (Dest1..Dest3 and Dummy Dest4)

$x_ij$ ; units shipped from plant i to destination j



Shipping cost from plant i to destination j = $S_ij$

production cost per unit at plant i = $P_i$

then, combined production cost $C_ij$ = $S_ij$ + $P_i$



B) Objective Function: 

minimizing total cost

Min TC = $305x_11$ + $306x_12$ + $312x_13$  
         + $203x_21$ + $205x_22$ + $210x_23$ 
         + $256x_31$ + $252x_32$ + $256x_33$ 
         + $153x_41$ + $154x_42$ + $160x_43$ 
         

C) Constraints 

1. supply constraints: 

plant 1: $x_11 + x_12 + x_13 + x_14 \le 100$

plant 2: $x_21 + x_22 + x_23 + x_24 \le 75$

plant 3: $x_31 + x_32 + x_33 + x_34 \le 85$

plant 4: $x_41 + x_42 + x_43 + x_44 \le 40$

2.Demand constraints: 

each destination has specific demand that needs to be fulfilled(including dummy destination demand)

$x_11 + x_21 + x_31 + x_41 \ge 80$

$x_12 + x_22 + x_32 + x_42 \ge 90$

$x_13 + x_23 + x_33 + x_43 \ge 70$

$x_14 + x_24 + x_34 + x_44 \ge 60$   (dummy)

Non-Negativity: 

All decisions must be non-negative since negative shipments are not possible.

$x_ij \ge 0$

Summary model: 

Min TC = $305x_11$ + $306x_12$ + $312x_13$ + $300x_14$ 
         + $203x_21$ + $205x_22$ + $210x_23$ + $200x_24$ 
         + $256x_31$ + $252x_32$ + $256x_33$ + $250x_34$
         + $153x_41$ + $154x_42$ + $160x_43$ + $150x_44$ 
         
subject to: 

supply constraints 

$x_11 + x_12 + x_13 + x_14 \le 100$

$x_21 + x_22 + x_23 + x_24 \le 75$

$x_31 + x_32 + x_33 + x_34 \le 85$

$x_41 + x_42 + x_43 + x_44 \le 40$

demand constraints 

$x_11 + x_21 + x_31 + x_41 \ge 80$

$x_12 + x_22 + x_32 + x_42 \ge 90$

$x_13 + x_23 + x_33 + x_43 \ge 70$

$x_14 + x_24 + x_34 + x_44 \ge 60$  

non-negativity: $x_ij \ge 0$


## Q2. SOLUTION IN R MARKDOWN 

```{r}
library(lpSolve)
```

setting objective function: 
```{r}
tc <- matrix(c(305, 306, 312, 0,
             203, 205, 210, 0,
             256, 252, 256, 0,
             153, 154, 160, 0), nrow = 4, byrow = TRUE)
```

setting constraints and RHS (supply side) 
```{r}
row.signs <- rep("<=", 4)
row.rhs <- c(100, 75, 85, 40)
```

setting constraints and RHS (demand side)
```{r}
col.signs <- rep(">=", 4)
col.rhs <- c(80, 90, 70, 60)
```


```{r}
# run the model 

lptrans <- lp.transport(tc, "min", row.signs, row.rhs, col.signs, col.rhs)

print(lptrans)
```

```{r}
# getting the values of 16 decision variables: 

lptrans$solution
```
```{r}
#Getting the value of the minimized cost 

lptrans$objval
```
when solved the transportation problem, the minimized value of transportation cost is: 55320


## Q3: DUAL FORMULATION 

A) Variables: 

let $u_i$ be the per unit cost at each plant i (1,2,..4) and $v_j$ be the demand at each destination j (1,2,..,4).

$c_ij$ - total combined cost of production including transportation for each unit at each destination j, for each demand. 


B) Objective function: 

the RHS of constraints of primal will be the co efficeints in dual. 

objective function is to maximize VA(value addition):

$Max VA = Revenue - cost$

Max Z = $80v_1 + 90v_2 + 70v_3 + 60v_4 - 100u_1 - 75u_2 - 85u_3 - 40u_4$ 

C) Constraints : 

$v_j - u_i \ge c_ij$  


plant 1 supplies to 4 possible destinations: 

eg. $v_1 - u_1 \ge c_11$ = 5 i.e $v_1\ge 5+300$, $u_1$ = 300   


$v_1 - u_1 \ge 305$ 

$v_2 - u_1 \ge 306$

$v_3 - u_1 \ge 312$


plant 2 supplies to 4 possible destinations:

$v_1 - u_2 \ge 203$

$v_2 - u_2 \ge 205$

$v_3 - u_2 \ge 210$


plant 3 supplies to 4 possible destinations:

$v_1 - u_3 \ge 256$

$v_2 - u_3 \ge 252$

$v_3 - u_3 \ge 256$


plant 4 supplies to 4 possible destinations:

$v_1 - u_4 \ge 153$

$v_2 - u_4 \ge 154$

$v_3 - u_4 \ge 160$


Non-Negativity 

$v_j \ge 0$ 

$u_i \ge 0$  

for j = 1,2,3,4 and i = 1,2,3. 



## Q4. ECONOMIC INTERPRETATION OF THE DUAL

1. Marginal cost and Marginal revenue 

$u_i$ → reflects MC(marginal cost): the cost of producing and supplying one more unit from plant i.

$v_i$ → reflects MR(marginal revenue): the benefit or price a destination j is willing to pay for receiving one more unit.

At equilibrium (optimal solution), for all active shipping routes (where $x_ij \gt 0$

in our constraint $v_j - u_i \ge c_ij$ 

which is $v_i \ge u_i +c_ij$

then, $v_1 \ge u_1 + 300$

plant 1 can keep on increasing production and shipping to destination 1 as long as  $v_1 \ge u_1 + 300$, which is $MR \GE MC$. ON the other hand, plant 1 reduces production and shipping if $v_1 \le u_1 + 300$ i.e., $MC \GE MR$, thats when production should stop. when MR = MC, thats the optimal point to stop producing. this is equilibrium for profit maximization. 

```{r}
library(ggplot2)
```

```{r}
quantities <- seq(0, 240, by = 1)
```

define MR(demand) and MC(supply) functions
```{r}
MR <- function(quantity) {
# Marginal Revenue (MR) or Demand Curve
150 - 1 * quantity # Decreasing as quantity increases
}
MC <- function(quantity) {
# Marginal Cost (MC) or Supply Curve
0.5 * quantity # Increasing as quantity increases
}
```

```{r}
# create data frame with quantity, MR and MC
data <- data.frame(
  quantity = quantities,
  MR = MR(quantities),
  MC = MC(quantities)
)
```

```{r}
# Example data
data <- data.frame(
  Quantity = seq(0, 200, by = 10),
  MR = 200 - 1 * seq(0, 200, by = 10),  # Downward-sloping MR curve
  MC = 50 + 1.5 * seq(0, 200, by = 10)  # Upward-sloping MC curve
)
```

Plot MR (Demand) and MC (Supply) curves with Price on y-axis and Quantity on x-axis
```{r}
# Plot MR and MC curves
library(ggplot2)

ggplot(data, aes(x = Quantity)) +
  geom_line(aes(y = MR, color = "MR (Marginal Revenue)"), size = 1) +     # MR curve
  geom_line(aes(y = MC, color = "MC (Marginal Cost)"), size = 1) +        # MC curve
  labs(
    title = "MR and MC Curves (Demand and Supply) with Intersection at Quantity ≈ 100",
    x = "Quantity",
    y = "Price"
  ) +
  scale_color_manual(
    values = c("MR (Marginal Revenue)" = "blue", "MC (Marginal Cost)" = "red")
  ) +
  theme_minimal() +
  theme(legend.title = element_blank())   # Remove legend title
```


Conditions for shipping, non shipping: 

1. Shipping occurs only when it is profitable in the equilibrium sense: $v_j - u_i \ge c_ij$ 

That is, the value of sending one more unit (destination’s valuation) exactly matches the cost (plant’s opportunity cost + transport) or creates profit.

2. shipping does not occur, If a route is not used in the optimal plan ($x_ij = 0$)

$v_j - u_i \lt c_ij$ 

The effective value of sending goods via that route is less than the cost so it’s not profitable to ship along that path.

3. ship, it carries flow, exactly break even point: 

$v_j - u_i = c_ij$




Arbitrage(hiring) and No Arbitrage(no hiring shipping company): 

if  $v_j - u_i \ge c_ij$ Arbitrage exists here, one could profit by directly shipping or hiring a shipping company until $v_j - u_i \le c_ij$ . 
there's no no upper bound and infinite production and supply.

However, in equilibrium:

The LP ensures no such route can exist (violates optimality).

If it did, the solution would reallocate shipments until the profit gap closes, restoring

   $v_j - u_i = c_ij$ 

Thus, no arbitrage can persist in an optimal transportation plan, the market self-adjusts to equilibrium.




## solving dual in r
```{r}
library(lpSolve)

costs <- matrix(c(305, 306, 312,
                203, 205, 210,
                256, 252, 256,
                153,154, 160), 
                nrow = 4, byrow = TRUE)
supply <- c(80,90,70,60)
demand <- c(100,75,85,40) 

m <- length(supply)
n <- length(demand)

#dual objective co efficients 
#order: u1..u2..
obj <- c(supply, demand)
```

```{r}
#Constraints: u_i + v_j <= c_ij
A <- matrix(0, nrow = m*n, ncol = m+n)
row <- 1
for (i in 1:m) {
for (j in 1:n) {
A[row, i] <- 1 # u_i
A[row, m+j] <- 1 # v_j
row <- row + 1
}
}
dir <- rep("<=", m*n)
b <- as.vector(t(costs)) # costs row-wise
# Solve dual
dual <- lp(direction = "max",
objective.in = obj,
const.mat = A, const.dir = dir, const.rhs = b)

```

```{r}
#results 
dual$status
```

```{r}
dual$solution
```
```{r}
dual$objval
```








